\chapter{The Curious Asymmetry}

The US government is currently about 40\% of GDP.\footnote{In international
comparisons, too often one sees the US Federal Government compared with total
government spending in other countries. This makes the US seem like a
small-government country. However, in the US Federal system, state and local
governments spend a large chunk of what government spending. It is also
noteworthy that many other countries manage their spending at an even more
local level, but that is rarely mentioned in international comparisons.}

If we look just at the economic role of government (which is the focus of this
book), we can place the possible governments from right to left. At one end of
the spectrum, we put Randian anarcho-capitalism and full-blown communism with
forced labor camps at the other. At the first end, individuals have all rights
and there are no societal rights; at the other end, individuals have no rights
and society owns them. Economically speaking, we are close to the middle of
this spectrum.\footnote{On non-economy dimensions, however, the 20th century
has seen a steady walk from the society to the individual end of the spectrum.
It is the most surprising political fact of the last 100 years that the
anarchists basically won on state interference on sexuality. The result may not
be the sexual utopia that radical socialists predicted nor the Randian promise
of a rational open sexuality (her own attempts were not as successful as those
of the character in her novels). However, in policy, we live in anarchy.
Despite the fact that much hurt comes from it and that there is widespread
condemnation of adultery, it is a matter that has become a private and social
matter, outside the boundaries of courts. This is, however, a topic for another
book.}

Consider the case of another basic right, that of free-speech. I could make the
following argument:

\begin{quote}
The idea that there is a free-speech right is an illusion. It might make us
feel good, but, upon further examination, we quickly detect the falsehood.
Historically, there has never been completely free speech. We cannot shout
``fire'' in a crowded movie theater and certain forms of obscenity have always
been banned.

Or consider the criminal master-mind who, upon getting arrested, claims that
``yes, he did say that so-and-so should be killed and even offered money for
it,'' but that is purely speech, he did not pull any triggers. Therefore, he
has a First Amendment Right to Free Speech and the charge of conspiracy is a
violation of his rights.

This is absurd. Society has an interest in what people say. Once you accept
this, you see that there is really no difference (except in degree) between
banning libel and banning criticism of elected officials.
\end{quote}

Naturally, we understand that the above argument is bunk. Accepting certain
restrictions to unbridled speech does not mean that there is no right to free
speech. It means that the free speech can only be regulated when equally
important rights are at stake. In institutional terms, this means that free
speech regulation has an extra layer of protection in the form of judicial
supervision that more run-of-the-mill legislation does not.

Valuing free speech does not mean that there are no limits to it, that does
limits cannot change in response to cultural changes (prohibitions on erotic
material are certainly much more relaxed than before), or that there can be
strong and vigorous disagreements on what is and is not an acceptable limit to
free speech (as the Citizens United ruling on campaign finance made clear,
there is not even complete agreement as to what is speech and what is not).

Yet, in the area of economic rights, too often, it seems that simply pointing
out that there are necessarily limits and have always been limits to economic
rights, means that the concept is meaningless. Instead, I think that we should
interpret the concept of economic rights like we interpret free speech:
violating those rights can be necessary, it can even be a good thing to do (if
it leads to a better society), but there should be a high level of scrutiny
and, unless there are strong arguments for the violation, we should preserve
people's rights.

Therefore, we can talk about economic rights and not necessarily defend Randian
anarcho-capitalism. Similarly, an argument that denies that these arguments are
absolute does not give the state full power to take someone's house and give it
to someone else because that person will pay more taxes.\footnote{In the US,
the government cannot take a single house, but it can take a whole neighborhood
and demolish it to give it to someone who will pay more taxes by declaring the
neighborhood as \emph{blighted}. Thus ruled the Supreme Court in \emph{Kelo vs.
City of New London}.}

\subsection{Curious Asymmetry Thoughts}

``Socialism versus anarcho-capitalism is a false dichotomy. Neither is a choice
in front of us right now.''

``I am not saying that property implies absolute rights, just that it implies
some rights and we have to be careful not to trampled them for little gain.''

``You have just argued forcefully for universal access to education, but there
is no logical link to any particular mode of providing that education.''

