\chapter{Introduction}

\epigraph{No man is a prophet in his own land}

The welfare state is one of the 20th century's achievements and I've come to
praise it, not bury it. In practice, however, the current welfare state suffers
from the vices of 20th century thinking. It is too trusting of bureacracies as
it assumes that there is a direct link between the goals of a bureacratic
machine and the outcomes.  It mistrusts the very individuals it is trying to
help. We need a \emph{better welfare state}.

Beneath all the rhetorical polarization in modern day America, there is central
ground on which both the Democrats and Republicans agree. The Democrats are not
socialists (they will not completely nationalize industry and hand it to
politically appointed technocrats). The Republicans will not eviscerate the
institutions of the welfare state (in fact, the last time that Republicans
controlled both the Congress and the Presidency, they used that power to expand
the welfare state, in the form of Medicare Part D). No matter who wins the next
election, the American government will be about as big as it is now (at most a
few percentage points bigger or smaller). Of course, there are differences,
there are important issues at stake, but that should not blind us to the middle
ground on which progress can be made.

The question is not whether to have a government or not, not whether the
government is the problem or the solution (it is rarely either, but it can
often be a hindrance or a help), nor even whether to have a welfare state or
not. The question we actually face is what kind of government to have. This
book will be part of this conversation.

There is a role for the state in guaranteeing everyone a set of basic goods so
that they may fully participate in society as equal citizens. This book will
not question this premise. However, the way in which state institutions perform
their tasks does not always respect the recipients of such aid. The usual
charge is that they breed dependency or complacency. I will also argue that
they insult its recipients. They also demean them by treating them as
recipients instead of empowering them.

\bigskip
\bigskip

The welfare state is a big topic. I will focus on a few subjects under that
heading: education, health care, basic income provisions, and the environment.
Environmental regulation is, strictly speaking, not a part of the welfare
system, but it fits the rest of the book.

I will discuss how American institutions currently operate and talk about
alternatives. These will not be utopian alternatives, but we will see examples
from both American states and local regions or other societies. In particular,
a few societies will keep coming up, the S-countries: Singapore, Sweden, and
Switzerland. Neither is a perfectly good model for the US to follow. Singapore
is not a very free country (although it is democratic). Sweden and Switzerland
are both free democracies, but are small countries, with a culturally and
racially homogeneous population (very different from the US). Still, we can
learn from their examples. Denmark and Finland, even though they do not start
with an S, will show up as well. The Nordic countries\footnote{I will keep
Norway out of this, not because there is anything wrong with its economic
system or culture, but because there is not much to be learned from an oil-rich
state.} are often darlings of the Left, but they could as well be darlings of
the free-market Right.

It may be the case that a complete rewrite of the system to a Universal Basic
Income as favoured, for example, by Charles Murray,\backnote{\emph{In Our
Hands: A Plan to Replace the Welfare State} by Charles Murray} is a much better
alternative. However, my focus here is, to borrow a phrase from the blog
\emph{Marginal Revolution}, ``small steps to a much better world.''

\bigskip
\bigskip

There are a few interlocked goals to this book. One is to save the economic
freedom arguments from those who have used them (and often abused them) from
the conservative side. For example, conservative activists will often use the
language of property rights to defend zoning laws that keep poor, dark-skinned,
people away from rich people. However, zoning laws are very much
anti-free-market.\footnote{See
\url{http://www.theatlantic.com/business/archive/2012/03/affordable-housing-and-social-engineering-in-new-jersey/255269/}
for a perfect example. Matt Yglesias makes the same point in his wonderful book
``The Rent is Too Damn High.''} % FIXME: inline the comment

The other is to present free-market solutions to social problems not only as
more efficient than current alternatives (which is how they are typically
defended), but \emph{as morally superior}. Not morally superior only because
they involve less coercion or better respect property rights (the
traditional libertarian goals), but morally superior because they imply higher
respect for the recipients. At the interface of the welfare state and the
recipients, there are no magic ``rights,'' but a real interaction between
recipients and bureaucracies. The alternative models I propose here will empower
the recipients. This in the literal sense that it will \emph{give power to them}.

A family who approaches a school system with a voucher that gives them choices
is more empowered and will get more respect than a family who, because they
cannot afford a private school, approaches their failing neighborhood school.

There is always almost complete choice for parents who can afford it. Well-off
parents can resort to private schools or simply buy a house in their desired
school district. The discussion is only whether that choice should be extended
to not-so-fortunate parents. We should not lose track of this simple fact.

In fact, if the current system was a voucher system, I do not think that there
would be very strong arguments against moving to a limited-choice system.
Currently, in the US, we can find systems ranging from voucher to assigned
school as well as many systems in between where there is a constrained amount of
choice (for example, magnet schools are a form of such constrained choice).
Therefore, we can look at experiences in different states and localities.

In Chapter~\ref{chpt:education}, I will discuss the empirical evidence on
school choice and whether children benefit from switching schools under that
system. I think that there is evidence for a small positive effect, but
\emph{even if there was no effect on grades or other outcomes, school choice
would be the more moral system because it treats recipients with greater
respect}.

In other chapters, I will reverse the focus and look at provision of services.
Too often the focus of welfare discussion is on the demand side. That is, how
to pay for it? Almost all of the discussion on health care reform in 2008--10
was about how to pay and who should pay, given a fixed set of resources. In the
health care chapter, we will look at the other side of the ledger, how to
provide services and I will argue that health care costs are so high, at least
in part, due to over-regulation in the provision of services. If it was easier
for doctors trained abroad (including Americans who attended medical schools in
other countries---an option that would become popular if it was available,
given American medical school costs) to practice in the US, if it was easier to
provide medical services in innovative ways; then health care could be cheaper
for everyone. But over-accreditalization (demanding that even subsidiary staff
have advanced degrees) and too many barriers to innovation have driven up
costs. Again, we need not see the alternatives as either no-regulation or our
current system. Every regulation has costs and benefits and the US has too
often taken the view that ``no cost is high enough.''

This is not a partisan book. I hope that both Democrats and Republicans can
learn something from it and see that the middle ground is larger than the
rhetoric lets on.

\bigskip
\bigskip
\bigskip

I wrote this book while living in three different countries. The ideas were
first developed in the United States and spring from its intelectual dialogue
(which exists not on television nor on radio, but in magazines and,
increasingly, the internet). This is why the book focuses on this society.

I actually started writing while living in Portugal, and I finished the book in
Luxembourg. These are all Western countries and share some relevant cultural
baggage. However, within the Western sphere, they are very different. Portugal
is "to the left" of all of both Luxembourg and the US, but is not actually able
to provide its citizens with any security as its promises are of little value.
In the US and Luxembourg, the state plays a smaller role (smaller in Luxembourg
than in the US). However, in Luxembourg, the state does a better job in
assuring security to its citizens.  In this, it is similar to a collection of
small Nordic states: Finnland, Denmark, Sweden, or Switzerland; or the more
distant Singapore and New Zealand.  All of these are very free-market, but with
a real safety net.

\begin{flushright}
Luis Pedro Coelho\\
Pittsburgh, Lisbon, and City of Luxembourg
\end{flushright}

